---
title: "HMM"
author: "Pat Healy"
date: "29 March 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Script for one training and one test HMM for hub data

##Read in data


```{r}
internal_df <- read.csv("internal.csv")
```
#Remove the columns of unused variables
```{r}
#df <- internal_df[ -c(11:12, 19:21) ]
```
##Take the first hub
```{r}
library(dplyr)
train_df <- filter(internal_df, HubRandomID == 15994)
```

```{r}
#train_df <- train_df[ -c(10:14, 17:21, 29:31) ]
```

#Turn activity types into binary variables
```{r}
#short names
unique(train_df$ShortName)
```
```{r}
#Adding new variable for how participants compare to average
train_df$TotalParticipants <- train_df$AdultParticipants + train_df$ChildParticipants
#removing outliers
train_df <- train_df[train_df$TotalParticipants <= 70,]
train_df <- train_df[train_df$TotalParticipants > 0,]
```
```{r}
train_df$AvgTotalParticipants <- mean(train_df$TotalParticipants)
```
```{r}
train_df$DifferenceInParticipants <- train_df$TotalParticipants - train_df$AvgTotalParticipants 
```


```{r}
#scale from 0 to 1
train_df$DifferenceInParticipants_scaled <- (train_df$DifferenceInParticipants-min(train_df$DifferenceInParticipants))/
  (max(train_df$DifferenceInParticipants)-min(train_df$DifferenceInParticipants))
```

```{r}
#round out the number for defined states
train_df <- train_df %>% rowwise %>%
  mutate(Part_state = ifelse(DifferenceInParticipants_scaled >= 0.223, 1,0))   #ifelse(DifferenceInParticipants_scaled <= 0.33, -1, 0)))
```



#Add in Month and day of the week
```{r}
library(lubridate)

library(tidyr)

train_df <- mutate(train_df, Month = month(dmy(ActivityDate)))
train_df <- mutate(train_df, Day = wday(dmy(ActivityDate)))
train_df <- mutate(train_df, Year = year(dmy(ActivityDate)))
```
###removing Jan as an outlier for low results

```{r}
train_df <- train_df[train_df$Month >= 2,]
#train_df <- train_df[train_df$Year >= 2019,]
```



```{r}
#Scale the days and months from 0 to 1
train_df$Month_Scaled <- (train_df$Month-min(train_df$Month))/
  (max(train_df$Month)-min(train_df$Month))

train_df$Day_Scaled <- (train_df$Day-min(train_df$Day))/
  (max(train_df$Day)-min(train_df$Day))
```



#Run the model
```{r}
# Variables fitted:
#"ShortName", "CategoryName", "Month_Scaled", "Day_Scaled"
 ds1.mod1.3s <- depmix(DifferenceInParticipants_scaled ~ 1,  
                       transition = ~ ShortName + Day_Scaled + Month_Scaled + CategoryName , data = train_df, nstates = 2, family=multinomial("identity"))
 ds1.fm1.3s <- fit(ds1.mod1.3s)
 
 summary(ds1.fm1.3s)
```


#Plotting the states
```{r}
library(ggplot2)
library(tidyverse)
#plotting Data Set 1 Training Model 1 with 3 states
probs = posterior(ds1.fm1.3s)
	
	
 	plot((probs$state), type='s', main='Implied States', xlab='', ylab='State')
 	
 
	matplot(probs[,-1], type='l', main='Probabilities', ylab='Probability')
		legend(x='topright', c('State1','State2'),  fill=1:2, bty='n')
		
		

```

#Using stepwise to find best order of variables for the model. Judgement is made based on how much they decrease the log likelihood. 

```{r}
#run through all variables to find highest log lik
#tm2.fcov1 is training model 2 fitted covariates 1
tm1.fcov1 <- lapply(train_df[c( "ShortName", "CategoryName", "Month_Scaled", "Day_Scaled" )], function(w) fit((depmix(DifferenceInParticipants_scaled ~ 1, data = train_df, transition = ~ w, nstates = 3, family=multinomial("identity"))), verbose = FALSE, emc=em.control(rand=TRUE)))
```

```{r}
#Pulling covergence log likelihoods into a dataframe
tm1.fcov1_df <-  as.data.frame(
  c(logLik(tm1.fcov1$ShortName), logLik(tm1.fcov1$CategoryName),
    logLik(tm1.fcov1$Month_Scaled), logLik(tm1.fcov1$Day_Scaled)
    )
  )
tm1.fcov1_df$newcolumn<-c( "ShortName", "CategoryName", "Month_Scaled", "Day_Scaled")
names(tm1.fcov1_df) <- c("convergence.loglik", "Variables")
#print variable with the highest convergence log likelihood
tm1.fcov1_df %>% 
   slice(which.max(convergence.loglik))
```
##Second best variable
```{r}
#run through all variables to find highest log lik
#tm2.fcov1 is training model 2 fitted covariates 1
tm1.fcov2 <- lapply(train_df[c( "CategoryName", "Month_Scaled", "Day_Scaled" )], function(w) fit((depmix(DifferenceInParticipants_scaled ~ 1, data = train_df, transition = ~ w, nstates = 3, family=multinomial("identity"))), verbose = FALSE, emc=em.control(rand=TRUE)))
```

```{r}
#Pulling covergence log likelihoods into a dataframe
tm1.fcov2_df <-  as.data.frame(
  c( logLik(tm1.fcov2$CategoryName),
    logLik(tm1.fcov2$Month_Scaled), logLik(tm1.fcov2$Day_Scaled)
    )
  )
tm1.fcov2_df$newcolumn<-c(  "CategoryName", "Month_Scaled", "Day_Scaled")
names(tm1.fcov2_df) <- c("convergence.loglik", "Variables")
#print variable with the highest convergence log likelihood
tm1.fcov2_df %>% 
   slice(which.max(convergence.loglik))
```

#find the last 2 variables

```{r}
#run through all variables to find highest log lik
#tm2.fcov1 is training model 2 fitted covariates 1
tm1.fcov3 <- lapply(train_df[c( "CategoryName", "Month_Scaled")], function(w) fit((depmix(DifferenceInParticipants_scaled ~ 1, data = train_df, transition = ~ w, nstates = 3, family=multinomial("identity"))), verbose = FALSE, emc=em.control(rand=TRUE)))
```

```{r}
#Pulling covergence log likelihoods into a dataframe
tm1.fcov3_df <-  as.data.frame(
  c( logLik(tm1.fcov2$CategoryName),
    logLik(tm1.fcov2$Month_Scaled)
    )
  )
tm1.fcov3_df$newcolumn<-c(  "CategoryName", "Month_Scaled")
names(tm1.fcov3_df) <- c("convergence.loglik", "Variables")
#print variable with the highest convergence log likelihood
tm1.fcov3_df %>% 
   slice(which.max(convergence.loglik))
```







We have something to compare to now so add holidays and funding info if available
Also remove outliers.



